The Spatial and Temporal Domains of Modern Ecology \\ OR Ecology's Spatio-temporal Domains\\ OR Space, Time, and Ecology

### 2016-04-07 Rmd/space-time.Rmd
Histogram figure added to main text as Figure 1 (ties into text describing primary statistics). Histogram figure and main kde figure (Fig. 2) sub-panel labels added.

# cxa = 0.7
ll <- 3.5
cxa = 0.8
# pdf("paper/figures/res_v_extent.pdf", width = 7, height = 2.5)
pdf("paper/figures/kde4.pdf", width = 7, height = 5.5)
par(mfrow = c(2, 2), mar = c(6, 6, 0.3, 3), oma = c(0, 0, 0.5, 1), 
    mgp = c(2, 0.25, 0))
# resolutions
# image(rres3, col = rev(terrain.colors(30)), breaks = brks,
image(rres3, col = rev(terrain.colors(50)), 
     axes = FALSE, xlab = "", ylab = "")
axis(1, at = taxis1$logdays[-c(5, 10)], labels = tlab1[-c(5, 10)], 
     las = 2, tcl = -0.2, cex.axis = cxa)
axis(2, at = aaxis1$logres[-1], labels = alab1[-1], las = 2, tcl = -0.2, 
     cex.axis = cxa)
mtext("Sampling interval", 1, cex = cxa, line = ll)
mtext("Spatial resolution", 2, cex = cxa, line = ll)
aargs <- list(mgp = c(3, 0.25, 0), at = brklabs, labels = brklabs, 
              cex.axis = cxa, tcl = -0.1)
plot(rres3, legend.only = TRUE, axis.args = aargs, legend.width = 1.25,
     #legend.shrink = 0.9, breaks = brks, col = rev(terrain.colors(30)))
     legend.shrink = 0.9, col = rev(terrain.colors(50)))

# extents
# image(rext3, col = rev(terrain.colors(30)), axes = FALSE, xlab = "", 
image(rext3, col = rev(terrain.colors(50)), axes = FALSE, xlab = "", 
      ylab = "")#, breaks = brks2)
axis(1, at = taxis2$logdays[-5], labels = tlab2[-5], las = 2,
     cex.axis = cxa, tcl = -0.2)
axis(2, at = aaxis2$logarea, labels = alab2, las = 2, cex.axis = cxa,
     tcl = -0.2)
aargs <- list(mgp = c(3, 0.25, 0), at = brklabs2, labels = brklabs2, 
              cex.axis = cxa, tcl = -0.1)
mtext("Temporal duration", 1, cex = cxa, line = ll)
mtext("Spatial extent", 2, cex = cxa, line = ll)
plot(rext3, legend.only = TRUE, axis.args = aargs, legend.width = 1.25,
     #legend.shrink = 0.9, breaks = brks2, col = rev(terrain.colors(30)))
     legend.shrink = 0.9, col = rev(terrain.colors(50)))

# spatial
image(sres, col = rev(terrain.colors(100)), axes = FALSE, xlab = "", ylab = "")
# points(spatdat$x, spatdat$y, cex = 0.1)
axis(1, at = aaxis1$logres, labels = alab1, las = 2, tcl = -0.2, cex.axis = cxa)
axis(2, at = aaxis2$logarea, labels = alab2, las = 2, tcl = -0.2, 
     cex.axis = cxa)
mtext("Spatial resolution", 1, cex = cxa, line = ll)
mtext("Spatial extent", 2, cex = cxa, line = ll)
aargs <- list(mgp = c(3, 0.25, 0), at = seq(0, 0.12, 0.02), 
              labels = seq(0, 0.12, 0.02), cex.axis = cxa, tcl = -0.1)
plot(sres, legend.only = TRUE, axis.args = aargs, legend.width = 1.25,
     legend.shrink = 0.9, col = rev(terrain.colors(100)))

# temporal
image(tres, col = rev(terrain.colors(50)), axes = FALSE, xlab = "", ylab = "")
# points(spatdat$x, spatdat$y, cex = 0.1)
axis(1, at = taxis1$logdays[-c(5, 10)], labels = tlab1[-c(5, 10)], 
     las = 2, tcl = -0.2, cex.axis = cxa)
axis(2, at = taxis2$logdays[-5], labels = tlab2[-5], las = 2,
     cex.axis = cxa, tcl = -0.2)
mtext("Sampling interval", 1, cex = cxa, line = ll)
mtext("Temporal duration", 2, cex = cxa, line = ll)
aargs <- list(mgp = c(3, 0.25, 0), at = brklabs, labels = brklabs, 
              cex.axis = cxa, tcl = -0.1)
plot(tres, legend.only = TRUE, axis.args = aargs, legend.width = 1.25,
     legend.shrink = 0.9, col = rev(terrain.colors(100)))
dev.off()

@@

pdf("paper/figures/hists.pdf", width = 7, height = 2.5)
par(mfrow = c(1, 4), mar = c(4, 2, 1, 1), oma = c(2, 3, 0, 0))
# resolution
h <- hist(hdat$res, breaks = aaxis1$logres, plot = FALSE)
h$density <- h$counts / sum(h$counts) * 100
plot(h, freq = FALSE, ylab = "", las = 2,  xaxt = "n", 
     xlab = "Spatial resolution", ylim = c(0, 25), main = "", col = "red1")
axis(1, at = aaxis1$logres, labels = alab1, las = 2, cex.axis = cxa, 
     tcl = -0.2, mgp = c(2, 0.25, -0.45))
mtext("Percent of observations", side = 2, line = 3, cex = 0.8)

# extent
h <- hist(hdat$ext, breaks = c(-6, aaxis2$logarea), plot = FALSE)
h$density <- h$counts / sum(h$counts) * 100
plot(h, freq = FALSE, ylab = "", las = 2,  xaxt = "n", 
     xlab = "Spatial extent", ylim = c(0, 20), main = "", col = "red3")
axis(1, at = c(-6, aaxis2$logarea), 
     labels = c(expression(paste(NULL<="0.01m"^2)), expression(paste(0.1, m^2)), 
                alab2[-1]), las = 2, cex.axis = cxa, 
     tcl = -0.2, mgp = c(2, 0.25, -0.45))

# interval
tax <- c(taxis2$logdays[-c(10:11)], mean(taxis2$logdays[c(10:11)]))
h <- hist(hdat$int, breaks = taxis1$logdays, plot = FALSE)
h$density <- h$counts / sum(h$counts) * 100
plot(h, freq = FALSE, ylab = "", las = 2,  xaxt = "n", 
     xlab = "Sampling interval", ylim = c(0, 30), main = "", col = "blue1")
axis(1, at = tax, labels = tlab1[-10], las = 2, cex.axis = cxa, 
     tcl = -0.2, mgp = c(2, 0.25, -0.45))

# duration
# tax <- c(taxis2$logdays[-c(10:11)], mean(taxis2$logdays[c(10:11)]))
h <- hist(hdat$dur, breaks = taxis2$logdays, plot = FALSE)
h$density <- h$counts / sum(h$counts) * 100
plot(h, freq = FALSE, ylab = "", las = 2,  xaxt = "n", 
     xlab = "Temporal duration", ylim = c(0, 30), main = "", col = "blue4")
axis(1, at = taxis2$logdays, labels = tlab2, las = 2, cex.axis = cxa, 
     tcl = -0.2, mgp = c(2, 0.25, -0.45))
dev.off()

>>>>

# cxa = 0.7
ll <- 3.5
cxa = 0.8
# pdf("paper/figures/res_v_extent.pdf", width = 7, height = 2.5)
pdf("paper/figures/kde4.pdf", width = 7, height = 5.5)
par(mfrow = c(2, 2), mar = c(6, 6, 0.3, 3), oma = c(0, 0, 0.5, 1), 
    mgp = c(2, 0.25, 0))
# resolutions
# image(rres3, col = rev(terrain.colors(30)), breaks = brks,
image(rres3, col = rev(terrain.colors(50)), 
     axes = FALSE, xlab = "", ylab = "")
axis(1, at = taxis1$logdays[-c(5, 10)], labels = tlab1[-c(5, 10)], 
     las = 2, tcl = -0.2, cex.axis = cxa)
axis(2, at = aaxis1$logres[-1], labels = alab1[-1], las = 2, tcl = -0.2, 
     cex.axis = cxa)
mtext("Sampling interval", 1, cex = cxa, line = ll)
mtext("Spatial resolution", 2, cex = cxa, line = ll)
aargs <- list(mgp = c(3, 0.25, 0), at = brklabs, labels = brklabs, 
              cex.axis = cxa, tcl = -0.1)
plot(rres3, legend.only = TRUE, axis.args = aargs, legend.width = 1.25,
     #legend.shrink = 0.9, breaks = brks, col = rev(terrain.colors(30)))
     legend.shrink = 0.9, col = rev(terrain.colors(50)))
mtext("A", side = 3, line = -1, cex = 0.8, adj = 0.05)

# extents
# image(rext3, col = rev(terrain.colors(30)), axes = FALSE, xlab = "", 
image(rext3, col = rev(terrain.colors(50)), axes = FALSE, xlab = "", 
      ylab = "")#, breaks = brks2)
axis(1, at = taxis2$logdays[-5], labels = tlab2[-5], las = 2,
     cex.axis = cxa, tcl = -0.2)
axis(2, at = aaxis2$logarea, labels = alab2, las = 2, cex.axis = cxa,
     tcl = -0.2)
aargs <- list(mgp = c(3, 0.25, 0), at = brklabs2, labels = brklabs2, 
              cex.axis = cxa, tcl = -0.1)
mtext("Temporal duration", 1, cex = cxa, line = ll)
mtext("Spatial extent", 2, cex = cxa, line = ll)
plot(rext3, legend.only = TRUE, axis.args = aargs, legend.width = 1.25,
     #legend.shrink = 0.9, breaks = brks2, col = rev(terrain.colors(30)))
     legend.shrink = 0.9, col = rev(terrain.colors(50)))
mtext("B", side = 3, line = -1, cex = 0.8, adj = 0.05)

# spatial
image(sres, col = rev(terrain.colors(100)), axes = FALSE, xlab = "", ylab = "")
# points(spatdat$x, spatdat$y, cex = 0.1)
axis(1, at = aaxis1$logres, labels = alab1, las = 2, tcl = -0.2, cex.axis = cxa)
axis(2, at = aaxis2$logarea, labels = alab2, las = 2, tcl = -0.2, 
     cex.axis = cxa)
mtext("Spatial resolution", 1, cex = cxa, line = ll)
mtext("Spatial extent", 2, cex = cxa, line = ll)
aargs <- list(mgp = c(3, 0.25, 0), at = seq(0, 0.12, 0.02), 
              labels = seq(0, 0.12, 0.02), cex.axis = cxa, tcl = -0.1)
plot(sres, legend.only = TRUE, axis.args = aargs, legend.width = 1.25,
     legend.shrink = 0.9, col = rev(terrain.colors(100)))
mtext("C", side = 3, line = -1, cex = 0.8, adj = 0.05)

# temporal
image(tres, col = rev(terrain.colors(50)), axes = FALSE, xlab = "", ylab = "")
# points(spatdat$x, spatdat$y, cex = 0.1)
axis(1, at = taxis1$logdays[-c(5, 10)], labels = tlab1[-c(5, 10)], 
     las = 2, tcl = -0.2, cex.axis = cxa)
axis(2, at = taxis2$logdays[-5], labels = tlab2[-5], las = 2,
     cex.axis = cxa, tcl = -0.2)
mtext("Sampling interval", 1, cex = cxa, line = ll)
mtext("Temporal duration", 2, cex = cxa, line = ll)
aargs <- list(mgp = c(3, 0.25, 0), at = brklabs, labels = brklabs, 
              cex.axis = cxa, tcl = -0.1)
plot(tres, legend.only = TRUE, axis.args = aargs, legend.width = 1.25,
     legend.shrink = 0.9, col = rev(terrain.colors(100)))
mtext("D", side = 3, line = -1, cex = 0.8, adj = 0.05)
dev.off()

@@

pdf("paper/figures/hists.pdf", width = 7, height = 2.5)
par(mfrow = c(1, 4), mar = c(4, 2, 1, 1), oma = c(2, 3, 0, 0))
# resolution
h <- hist(hdat$res, breaks = aaxis1$logres, plot = FALSE)
h$density <- h$counts / sum(h$counts) * 100
plot(h, freq = FALSE, ylab = "", las = 2,  xaxt = "n", 
     xlab = "Spatial resolution", ylim = c(0, 25), main = "", col = "red1")
axis(1, at = aaxis1$logres, labels = alab1, las = 2, cex.axis = cxa, 
     tcl = -0.2, mgp = c(2, 0.25, -0.45))
mtext("Percent of observations", side = 2, line = 3, cex = 0.8)
mtext("A", side = 3, line = -1.5, cex = 0.8, adj = 0.05)

# extent
h <- hist(hdat$ext, breaks = c(-6, aaxis2$logarea), plot = FALSE)
h$density <- h$counts / sum(h$counts) * 100
plot(h, freq = FALSE, ylab = "", las = 2,  xaxt = "n", 
     xlab = "Spatial extent", ylim = c(0, 20), main = "", col = "red3")
axis(1, at = c(-6, aaxis2$logarea), 
     labels = c(expression(paste(NULL<="0.01m"^2)), expression(paste(0.1, m^2)), 
                alab2[-1]), las = 2, cex.axis = cxa, 
     tcl = -0.2, mgp = c(2, 0.25, -0.45))
mtext("B", side = 3, line = -1.5, cex = 0.8, adj = 0.05)

# interval
tax <- c(taxis2$logdays[-c(10:11)], mean(taxis2$logdays[c(10:11)]))
h <- hist(hdat$int, breaks = taxis1$logdays, plot = FALSE)
h$density <- h$counts / sum(h$counts) * 100
plot(h, freq = FALSE, ylab = "", las = 2,  xaxt = "n", 
     xlab = "Sampling interval", ylim = c(0, 30), main = "", col = "blue1")
axis(1, at = tax, labels = tlab1[-10], las = 2, cex.axis = cxa, 
     tcl = -0.2, mgp = c(2, 0.25, -0.45))
mtext("C", side = 3, line = -1.5, cex = 0.8, adj = 0.05)

# duration
# tax <- c(taxis2$logdays[-c(10:11)], mean(taxis2$logdays[c(10:11)]))
h <- hist(hdat$dur, breaks = taxis2$logdays, plot = FALSE)
h$density <- h$counts / sum(h$counts) * 100
plot(h, freq = FALSE, ylab = "", las = 2,  xaxt = "n", 
     xlab = "Temporal duration", ylim = c(0, 30), main = "", col = "blue4")
axis(1, at = taxis2$logdays, labels = tlab2, las = 2, cex.axis = cxa, 
     tcl = -0.2, mgp = c(2, 0.25, -0.45))
mtext("D", side = 3, line = -1.5, cex = 0.8, adj = 0.05)

dev.off()

### 2016-04-07 paper/main.tex
Histogram figure added to main text as Figure 1 (ties into text describing primary statistics). Histogram figure and main kde figure (Fig. 2) sub-panel labels added.

%16 + 29 + 26 + 67 + 14 + 37 + 10
In terms of spatial resolution, here defined as the two-dimensional space in which all measurable features of a natural object were recorded (as opposed to sub-sampled), the majority 63\% were collected in plots having resolutions $<$1 m$^2$, while 25\% were collected within plots of 1 m$^2$, up to 1 ha, and the remaining 12\% in plots of $\geq$1 ha. The total spatial extent covered (the number of sampled sites multiplied by the spatial resolution) by 85\% of observations was $<$10 ha, while 31\% covered less than 1 m$^2$.  Only 7\% covered an extent $\geq$10 ha, with just 1.1\% spanning areas $\geq$10 million ha.

In the temporal dimensions, 30\% of the assessed observations were ``once-offs'' that were not repeated. High frequency observations (ranging between as or more frequent than once per second up to daily) comprised 25\% of observations, 20\% were made at daily to monthly time steps, while 35\% and 4\% were respectively made at monthly up to yearly and yearly to decadal intervals.  The temporal duration of studies--the total amount of time the ecological feature was observed (the number of repeat observations multiplied by the effective sampling duration, SI)--was less than 1 day for 59\% of sampled observations,  35\% between 1 day and 1 year, and just 6\% covering greater than 1 year (including several paleoecological studies covering centuries).

Relationships between scales (Figure 1)
Vast majority of cases, plot resolution scales directly with the total extent of the study--these are field studies.  Those with

@@

\begin{enumerate}
 \item Intro/rationale 
% \begin{enumerate}
%   \item Understanding of natural patterns and processes depends upon the scales at which they are observed \cite{levin_problem_1992}(maybe   provide an example or two, could just come from Schneider)
%    \item Scale is an increasingly measured aspect of ecosystems \cite{schneider_rise_2001}
%    \item Given this importance, and the increasing awareness of scale in ecology, it is timely to consider the scales at which ecosystems are observed, as this provides insight into how our understanding of ecosystems might be affected (work in Levin's language about this). 
%    \end{enumerate}
   \item{What we did}
   \begin{enumerate}
%    \item Examined representative sample of papers from top 30 ecology journals--random sample of all titles from 2005-2014. Selected studies having observation component, excluding experimental manipulations, purely theoretical studies, commentaries, etc. 
%    \item Extracted information on the spatial resolution of the study, that is, the two-dimensional space that was observed in the study; the temporal analog, which was the interval between repeat observations; the total spatial extent covered by the observations; the total time period covered. 
    \item We recorded the scales that were actually covered by the observations, rather than potential scales that the collected data might represent--this would require information on autocorrelation length, which is not provided. 
    \subitem For spatial resolution, this was the finest unit representing complete spatial coverage, etc. etc. 
    \subitem Limitation for ocean studies, we did not study third dimension, volume (might be more relevant). 
   \end{enumerate}
   \item{What we found} 
   \begin{enumerate}
%    \item We reviewed a total of X randomly selected papers, of which 148 met our inclusion criteria and from which 367 distinct records related to observations made were collected. [Say something about size of sample and amount of time collected???].  
    \item  Of these, the vast majority (X\%) were field-based studies, while Y\% involved automated collection of data via instrumentation, and just Z\% made use of remote sensing.   
    \item From a perspective of temporal resolution, the majority of observations made in ecology are not repeated.
    \item Spatial resolution - most observations are between 10cm$^2$ to 10m$^2$
    \item Spatial extent
    \item Temporal extent
   \end{enumerate}
   \item{Implications} 
   \begin{enumerate}
    \item Although limited sample size, insight into the scales at which science of ecology is making bulk of observations
    \item Despite recognized importance of making observation across scales \cite{levin_problem_1992}, this does not appear to be happening. 
    \item Technical capabilities for making observations are not achieving adoption, despite potential of remote sensing to provide multi-scale ecological data \cite{estes_predictive_2011,estes_habitat_2008} 
   \end{enumerate}

>>>>

%16 + 29 + 26 + 67 + 14 + 37 + 10
In terms of spatial resolution, here defined as the two-dimensional space in which all measurable features of a natural object were recorded (as opposed to sub-sampled), the majority 63\% were collected in plots having resolutions $<$1 m$^2$, while 25\% were collected within plots of 1 m$^2$ up to 1 ha, and the remaining 12\% in plots of $\geq$1 ha (Fig. 1A). The total spatial extent covered (the number of sampled sites multiplied by the spatial resolution) by 85\% of observations was $<$10 ha, while 31\% covered less than 1 m$^2$ (Fig. 1B).  Only 7\% covered an extent $\geq$10 ha, with just 1.1\% spanning areas $\geq$10 million ha.

\begin{figure}[!ht]
%\begin{wrapfigure}{c}{1\textwidth}
\includegraphics[width=1\textwidth]{figures/hists.pdf}
\vspace{-0.15 cm}
\caption{Histograms of the spatial resolution (A) and extent (B), sampling interval (C) and temporal duration (D) of ecological observations collected from the surveyed ecological studies. Bars represent percentage of the 367 collected records falling within each bin.}
\label{afoto1}
\end{figure}

In the temporal dimensions, 30\% of the assessed observations were ``once-offs'' that were not repeated (Fig. 1C). High frequency observations (ranging between as or more frequent than once per second up to daily) comprised 25\% of observations, 20\% were made at daily to monthly time steps, while 35\% and 4\% were respectively made at monthly up to yearly and yearly to decadal intervals.  The temporal duration of studies--the total amount of time the ecological feature was observed (the number of repeat observations multiplied by the effective sampling duration, SI)--was less than 1 day for 59\% of sampled observations,  35\% between 1 day and 1 year, and just 6\% covering greater than 1 year (including several paleoecological studies covering centuries to millenia; Fig. 1D).

The juxtaposition of these scales of observations provides further insight into the spatio-temporal distribution of ecological observations (Fig. 2).  
Vast majority of cases, plot resolution scales directly with the total extent of the study--these are field studies.  Those with

@@

\begin{enumerate}
% \item Intro/rationale 
%    \begin{enumerate}
%   \item Understanding of natural patterns and processes depends upon the scales at which they are observed \cite{levin_problem_1992}(maybe   provide an example or two, could just come from Schneider)
%    \item Scale is an increasingly measured aspect of ecosystems \cite{schneider_rise_2001}
%    \item Given this importance, and the increasing awareness of scale in ecology, it is timely to consider the scales at which ecosystems are observed, as this provides insight into how our understanding of ecosystems might be affected (work in Levin's language about this). 
%    \end{enumerate}
   \item{What we did}
   \begin{enumerate}
%    \item Examined representative sample of papers from top 30 ecology journals--random sample of all titles from 2005-2014. Selected studies having observation component, excluding experimental manipulations, purely theoretical studies, commentaries, etc. 
%    \item Extracted information on the spatial resolution of the study, that is, the two-dimensional space that was observed in the study; the temporal analog, which was the interval between repeat observations; the total spatial extent covered by the observations; the total time period covered. 
    \item We recorded the scales that were actually covered by the observations, rather than potential scales that the collected data might represent--this would require information on autocorrelation length, which is not provided. 
%    \subitem For spatial resolution, this was the finest unit representing complete spatial coverage, etc. etc. 
    \subitem Limitation for ocean studies, we did not study third dimension, volume (might be more relevant). 
   \end{enumerate}
   \item{What we found} 
   \begin{enumerate}
%    \item We reviewed a total of X randomly selected papers, of which 148 met our inclusion criteria and from which 367 distinct records related to observations made were collected. [Say something about size of sample and amount of time collected???].  
    \item  Of these, the vast majority (X\%) were field-based studies, while Y\% involved automated collection of data via instrumentation, and just Z\% made use of remote sensing.   
%    \item From a perspective of temporal resolution, the majority of observations made in ecology are not repeated.
%    \item Spatial resolution - most observations are between 10cm$^2$ to 10m$^2$
%    \item Spatial extent
%    \item Temporal extent
   \end{enumerate}
   \item{Implications} 
   \begin{enumerate}
    \item Although limited sample size, insight into the scales at which science of ecology is making bulk of observations
    \item Despite recognized importance of making observation across scales \cite{levin_problem_1992}, this does not appear to be happening. 
    \item Technical capabilities for making observations are not achieving adoption, despite potential of remote sensing to provide multi-scale ecological data \cite{estes_predictive_2011,estes_habitat_2008} 
   \end{enumerate}
